Statistical quantification of confounding bias in predictive modelling

### 2021-10-29 manuscript.tex
Updates from Overleaf

% keywords can be removed
\keywords{machine learning, predictive modelling, confounder test, conditional independence, conditional permutation}

@@

Powerful and robust statistical tests for quantifying confounding bias in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches. Although it is tempting to think about confounding bias as the \emph{conditional dependence} of the model predictions on the confounder, given the target variable, the proper evaluation of the conditional independence among these variables is challenging. Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation) are not valid measures of conditional independence. Although warnings about this issue were given from early on\citep{korn1984ranges}, and received a fair amount of attention recently\citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g. feature-set characteristics and model regularization\citep{garcia2009study, kristensen2017whole} (see Supplementary Material \ref{sup:nomlinviol} for a simplistic example).

@@

@@

% fig: overview
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/overview.eps}
  \caption{\textbf{Overview of the proposed partial confounder test}. \\
    The partial confounder test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM). The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m). The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confounder variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. Figure source code available as jupyter notebook: \href{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}
  }
  \label{fig:overview}
\end{figure}

@@

%fig:sim-h0-demo
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row). Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1) and dependency of the mean is depicted with blue dotted lines an all plots. False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values. The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065. Values below this threshold indicate a valid type I error control and colored white.
  }
  \label{fig:sim-h0-demo}
\end{figure}

@@

@@

%fig:sim-normal
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type I error control and power of the partial confounder test based on simulations with normal conditional distribution and linear dependencies.} \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns). Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%. First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounding bias).
  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher 95\% confidence limit for a positive rate of $alpha=0.05$ is 0.11 for each tile.
  }
  \label{fig:sim-normal}
\end{figure}

@@

As suggested by theory (see \hyperref[sec:methods]{Methods} for details) and shown by the simulations with a wide range of settings, both of the proposed tests provide a valid Type I error control (Fig. \ref{fig:sim-normal}), even in case of non-linearity and non-normality (Figs. \ref{fig:sim-h0-demo} and \ref{fig:sim-non-normal}), except when non-normality is very extreme (purple distribution on Fig. \ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6).

@@

Estimates of statistical power for the partial confounder test (with normal and linear simulations, for  a wide range of parameters) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%. With a sample size of 500, the same confounding bias is still detected with a power greater than 84\% in all of the simulation cases. A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%. Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 25\% of variance, it is almost certainly detected even with the low sample size of $n \geq 50$.

Simulations with non-normal and non-linear dependencies has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}).
Type I error and power was found to be highly similar in case of categorical variables and for the \emph{full} confounder test, as well (Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full} and \ref{fig:sim-bbb-sig-full}).

@@

To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, they have been deployed in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
  \caption{\textbf{Acquisition- and age-bias of fluid intelligence prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data. This bias is robustly detected by the partial confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is given on the graphs via $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
  Solid red line between the confounder and the prediction means significant confounding bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the partial confound test. P-values of the 'full' confound test were all less then 0.0001, i.e. the confounders did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

Functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig. \ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.

After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence of confounding bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig. \ref{fig:hcp}), neither for acquisition batch nor for age. Both feature regression and COMBAT increased the predictive performance in the case of both confounders, with COMBAT providing best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.png}
  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups. In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX). AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD). The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001). While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias). COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Solid red line between the confounder and the prediction means significant confounding bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the partial confound test. P-values of the full confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:abide}
\end{figure}

Functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig. \ref{fig:abide}). Several sites (e.g. Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately associated ($R^2=0.169$ and $0.111$, respectively) to a lower probability for the participant to be assigned to the ASD group by the model.

Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.009$ for feature regression and COMBAT, respectively (second and third columns of the first row on Fig \ref{fig:abide}).

@@

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & \textbf{<0.0001} & <0.0001 \\
    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.73 & <0.0001 \\
    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.65 & <0.0001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & \textbf{<0.0001} & <0.0001 \\
    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.2 & <0.0001 \\
    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.16 & <0.0001 \\
\hline
ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & \textbf{<0.0001} & <0.0001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & \textbf{0.04} & <0.0001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & \textbf{0.009} & <0.0001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & \textbf{<0.0001} & <0.0001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & \textbf{<0.0001} & 0.1 \\
        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.59 & <0.0001 \\

\end{tabular}
\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated confounder-mitigation methods. Significant confounding bias is denoted by bold font.}
\end{table}

@@

The assessment of confounding bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings. However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig. \ref{fig:sim-h0-demo}). While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e. the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confounding bias testing must be designed so that its suitability for the particular problem may be judged easily.

In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confounder problem' in predictive modelling. Namely, the proposed approach places assumptions only about the conditional distribution of the confounder and the target variables, as required by the applied generalized additive\cite{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}. In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analysis, as well. Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and data acquisition is probably already designed so that the assumptions are likely fulfilled.
Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization)\citep{garcia2009study, kristensen2017whole}. 
This property distinguishes the approach from other alternatives as it guarantees a valid type-I error control even in cases of non-normally and non-linearly dependent predictions, i.e. in cases where Pearson and Spearman partial correlations, and many other methods fail.

@@

As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except extreme non-normality), confirming the tests' can be deployed in most predictive modelling scenarios. While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculation in order to identify the necessary sample size for proper model diagnostics.
Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.

%In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.\citep{spisak2020pain}. However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable. In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association. 
%Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
%These examples clearly indicate that guiding model diagnostics based on unconditioned associations between the confounder and the model output might result in overly strict decisions. More realistic is to aim at models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e. the prediction-confounder association is proportional to, and can be fully explained by, the target-confounder association.
%Put simply, this is what is tested by the proposed partial confounder test.

>>>>

% keywords can be removed
\keywords{machine learning, predictive modelling, confoudner bias, confounder test, conditional independence, conditional permutation}

@@

Powerful and robust statistical tests for quantifying confounding bias in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches. Although it is tempting to think about confounding bias as the \emph{conditional dependence} of the model predictions on the confounder, given the target variable, the proper evaluation of the conditional independence among these variables is challenging. Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation, see Fig. \ref{fig:sim-h0-demo}) are not valid measures of conditional independence. Although warnings about this issue were given from early on\citep{korn1984ranges}, and received a fair amount of attention recently\citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g. feature-set characteristics and model regularization\citep{garcia2009study, kristensen2017whole} (see Supplementary Material \ref{sup:nomlinviol} for a simplistic example).

@@

%fig:sim-h0-demo
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.5\paperwidth]{fig/sim_h0_demo.eps}
  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row). Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1). False positive rates for various simulation settings is shown in heatmaps. The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065. Values below this threshold indicate a valid type I error control and colored white.
  }
  \label{fig:sim-h0-demo}
\end{figure}

@@

% fig: overview
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.6\paperwidth]{fig/overview.eps}
  \caption{\textbf{Overview of the proposed partial confounder test}. \\
    The partial confounder test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM). The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m). The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confounder variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. Figure source code available as jupyter notebook: \href{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}
  }
  \label{fig:overview}
\end{figure}

@@

@@

%fig:sim-normal
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.6\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type I error control and power of the partial confounder test based on simulations with normal conditional distribution and linear dependencies.} \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns). Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%. First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounding bias).
  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher 95\% confidence limit for a positive rate of $alpha=0.05$ is 0.11 for each tile.
  }
  \label{fig:sim-normal}
\end{figure}

@@

@@

As suggested by theory (see \hyperref[sec:methods]{Methods} \ref{sec:methods-pct} for details) and shown by the simulations with a wide range of settings, both of the proposed tests provide a valid Type I error control (Fig. \ref{fig:sim-normal} and Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-ccc-lin-full} and \ref{fig:sim-bbb-lin-full}), even in case of non-linearity and non-normality (Figs. \ref{fig:sim-h0-demo}, \ref{fig:sim-non-normal} and Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-sig-full}), except when non-normality is very extreme (purple distribution on Fig. \ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6).

@@

Estimates of statistical power for the partial confounder test (with normal and linear simulations, for  a wide range of parameters) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing only $\sim 4\%$ to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%. With a sample size of 500, the same confounding bias is still detected with a power greater than 84-100\% in all of the simulation cases. A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) to achieve a reasonable level of power (75-98\%). Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 25\% of variance, it is almost certainly detected even with a low sample size of $n \geq 50$.

Simulations show that non-normality has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}). Simulations with sigmoid dependence resulted in an apparent loss of statistical power, however this is simply a consequence of the simulation methodology: with the same parameters, the sigmoid transformed confounder explains only approximately half the variance as compared to linear simulations.
Type I error control was valid in case of categorical variables and for the \emph{full} confounder test, as well. Power in these cases was also highly similar to the partial confoudner test results with numerical variables. (Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-partial} and \ref{fig:sim-bbb-sig-full}).

@@

To demonstrate the usefulness of the proposed tests in detecting various types of confounding bias, they have been deployed in two typical research scenarios - a regression and a classification problem - where confounder effects are known to hamper the development of biomedically valid predictive models.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
  \caption{\textbf{Acquisition batch and age bias of fluid intelligence prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder bias both for acquisition and age in the models trained on the raw data. This bias is robustly detected by the partial confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($Gf$) and predicted ($\hat{y}$) intelligence scores and the confounder variables is given on the graphs via $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improve the predictive performance.
  Solid red line between the confounder and the prediction means significant confounding bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the partial confounder test. P-values of the 'full' confounder test (not shown) were all less then 0.0001, i.e. the confounders did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

Functional connectivity data from the Human Connectome Project\citep{van2013wu} (HCP) was used to build predictive models of fluid intelligence ($G_f$) and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).

The partial confounder test revealed that the 'raw' model (without confounder mitigation) was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig. \ref{fig:hcp}) with later phases of the acquisition and lower age being associated to larger predicted values.

After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence for confounding bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig. \ref{fig:hcp}), neither for acquisition batch nor for age. Both feature regression and COMBAT increased the predictive performance, with COMBAT providing the overall best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.png}
  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups. In the top panel, predictions are plotted for each center separately. Color indicates the true diagnosis (DX). AT the bottom plot, color indicates the normalized index of in-scanner motion (normalized FD). The proposed confounder test reveals significant center and motion bias in the model trained on the raw data (p<0.0001). While feature regression seems to effectively mitigate center bias, it fails to remove motion bias (actually introduces a paradoxical negative bias). COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
  Relation between the true ($\hat{y}$) and predicted diagnosis scores and the confounder variables is shown by the graphs as $R^2$ values. Solid red line between the confounder and the prediction means significant confounding bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the partial confound test. P-values of the full confound test (not shown) were all less then 0.0001, i.e. the confounders did not fully drive the prediction for any of the models.
  }
  \label{fig:abide}
\end{figure}

Functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion and center bias (as previously reported e.g. in refs\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).

The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig. \ref{fig:abide}). Predictions for several sites (e.g. Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) were severely miscalibrated and higher motion was associated to a higher probability for ASD diagnosis.

Both feature regression and COMBAT seemed to significantly attenuate center bias, however the partial confound test still provided evidence for a significant bias ($p=0.04$ and $0.009$ for feature regression and COMBAT, respectively (second and third columns of the first row on Fig \ref{fig:abide}).

@@

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & \textbf{<0.0001} & <0.0001 \\
    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.73 & <0.0001 \\
    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.65 & <0.0001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & \textbf{<0.0001} & <0.0001 \\
    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.2 & <0.0001 \\
    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.16 & <0.0001 \\
\hline
ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & \textbf{<0.0001} & <0.0001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & \textbf{0.04} & <0.0001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & \textbf{0.009} & <0.0001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & \textbf{<0.0001} & <0.0001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & \textbf{<0.0001} & \textbf{0.1} \\
        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.59 & <0.0001 \\

\end{tabular}
\caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated confounders (conf.) confounder-mitigation methods (method). Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confoudner, as identified by the full confounder test.}
\end{table}

@@

The assessment of confounding bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings. However, statistical testing of conditional independence is challenging: 'conditional analogs' of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig. \ref{fig:sim-h0-demo}). While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e. the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confounding bias testing must be designed so that its suitability for the particular problem may be judged easily.

In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confounder problem' in predictive modelling. Namely, the proposed approach places assumptions only about the conditional distribution of the confounder and the target variables, as required by the applied generalized additive\cite{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}. In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analyses, as well. Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and the data acquisition is probably already designed so that the assumptions are likely fulfilled.
Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization)\citep{garcia2009study, kristensen2017whole}. 
This property distinguishes the approach from other alternatives as it guarantees a valid type-I error control even in cases of non-normally and non-linearly dependent predictions, i.e. in cases where Pearson and Spearman partial correlations, and many other methods fail.

@@

As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except extreme non-normality), confirming that the tests can be deployed in a wide variety of predictive modelling scenarios. While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculation in order to identify the necessary sample size for proper model diagnostics.

A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making\citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders typical for biomedical research (e.g. sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g. imaging artifacts) presents a great challenge to these efforts.
The usefulness of the proposed tests is demonstrated in two such examples, using the HCP\citep{van2013wu}  and the ABIDE\citep{di2014autism} datasets.

In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting} and could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model for acquisition batch can also be problematic, especially as it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed bias. Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course).

In the ABIDE dataset, neither the center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD. For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion (as known for many conditions, among others ADHD\citep{eloyan2012automated} or Alzheimer's disease \citep{rao2017predictive}).
